Literate statistics programming course - Assignment 1
======================================

1. Loading and preprocessing the data
--------------
```{r}
activity <- read.csv("activity.csv")
```

2. What is mean total number of steps taken per day?
--------------
```{r fig.width=7, fig.height=6}
total <- aggregate(activity$steps, by = list(activity$date), sum)
colnames(total) <- c("Day", "Steps")
hist(total$Steps, breaks = 20, main = "Total steps a day", xlab = "Steps per day")
totalMean <- mean(total$Steps, na.rm = T)
totalMedian <- median(total$Steps, na.rm = T)
options(scipen=999, digits=2) #Convenient output
```
Total number of steps taken per day median is `r totalMedian` and mean `r totalMean`.

3. What is the average daily activity pattern?
---------------
``` {r}
dayMean <- aggregate(activity$steps, by = list(activity$interval), function(x) mean(x, na.rm = TRUE))
colnames(dayMean) <- c("Time_interval", "Mean_steps")
library(ggplot2)
ggplot(aes(x = Time_interval, y = Mean_steps), data = dayMean, group = 1) + 
  geom_line() +
  labs(title = "Daily steps pattern")
maxPointer <- which(dayMean$Mean_steps == max(dayMean$Mean_steps))
maxMean <- dayMean[maxPointer,"Mean_steps"]
maxInt <- dayMean[maxPointer,"Time_interval"]

```
The highest average steps activity `r maxMean` is in the `r maxInt`th interval.

4. Imputing missing values
----------

```{r }
missing <- sum(is.na(activity$steps))
```
The activity dataset contains `r missing` missing values.
Therefore, the missing values were substituted with average day activity for the particular interval.

```{r}
#create a reference step object for the two months period as if there was average activity in every time interval
#find missing values in the original datase and use the same index to find values to impute
refStep <- rep(dayMean$Mean_steps, 61)
toSubstitute <- which(is.na(activity$steps))
activity2 <- activity
activity2$steps[toSubstitute] <- refStep[toSubstitute]
#characteristics of the missing values
substituted <- median(refStep[toSubstitute])

```{r fig.width=7, fig.height=6}
total2 <- aggregate(activity2$steps, by = list(activity2$date), sum)
colnames(total2) <- c("Day", "Steps")
hist(total2$Steps, breaks = 20, main = "Total steps a day", xlab = "Steps per day")
totalMean2 <- mean(total2$Steps, na.rm = T)
totalMedian2 <- median(total2$Steps, na.rm = T)
```

The daily steps median of the activity dataset with fixed NAs is `r totalMedian2` and mean `r totalMean2`. Median was robust against the NA values in the original dataset, mean shifted downwards in the corrected dataset. This is likely because the imputed values, which had to be ommited in the original dataset, were low, their median was `r substituted`. 

5. Are there differences in activity patterns between weekdays and weekends?
------------

```{r}
days <- weekdays(as.Date(activity2$date))
days[!(days %in% c("Saturday", "Sunday"))] <- "weekday"
days[days %in% c("Saturday", "Sunday")] <- "weekend"
activity2$day <- days
total2 <- aggregate(activity2[,c("steps", "interval")], by = list(activity2$interval, 
                                         "Days" = activity2$day), mean)

library(lattice)
xyplot(steps ~ interval|Days, data = total2, type = "l", layout=c(1,2))
```
Average patterns of the weekdays and weekend days are different. There is a higher activity around time interval 

